---
title: "Exploring IVF data"
author: "Marina Vabistsevits"
date: "27/03/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning =  FALSE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(comment = "")
```

```{r}
library(tidyverse)
library(vroom)
library(cowplot)
library(wesanderson)
library(scales)
library(waffle)
library(hrbrthemes)
library(patchwork)
library(yarrr)
```


```{r}
#rdsf
data_path<-"/projects/MRC-IEU/research/projects/ieu2/p4/052/working/data/hfea-ivf/data/"
# local 
data_path<-"~/Downloads/"
```

#### Load data 
```{r}
raw_data <-vroom(paste0(data_path, "hfea-ivf.csv"))
full_data <-raw_data %>% 
            # tidy age groups
            mutate(`Patient Age at Treatment`= gsub(" ", "", `Patient Age at Treatment`) ) %>% 
            # update year cols
            mutate(years=ifelse(source == "data/raw_data/ar-2010-2014-xlsb", '2010-2014',
                         ifelse(source == "data/raw_data/ar-1995-1999-xlsb", '1995-1999',
                         ifelse(source == "data/raw_data/ar-2000-2004-xlsb", '2000-2004',
                         ifelse(source == "data/raw_data/ar-1991-1994-xlsb", '1991-1994',
                         ifelse(source == "data/raw_data/ar-2015-2016-xlsb", '2015-2016',
                         ifelse(source == "data/raw_data/ar-2005-2009-xlsb", '2005-2009', 
                         NA))))))) 
                
dim(full_data)
full_data %>% count(years) %>% arrange(years)
```

```{r}
colnames(full_data)
```

## Plan

I'm going to explore the following categories:

* reasons for enrolment
* types of infertility
* cause of infertility
* types of IVF treatement
  - subtypes
  - how treatment types preference changed over time
  



### Exploring reasons for enrolment into IVF Registry
```{r}
# check partipation reasons
full_data %>% 
  count(`Main Reason for Producing Embroys Storing Eggs`)

# tidy up reason, keeping just the first one
reasons<-full_data %>% 
  count(`Main Reason for Producing Embroys Storing Eggs`) %>% 
  separate(1, into=c("main_reason"),sep=",") %>% 
  mutate(main_reason=str_trim(main_reason, side="right") )%>%  
  mutate(main_reason=ifelse(is.na(main_reason), "Reason unassigned", main_reason)) %>% 
  group_by(main_reason) %>% 
  summarize(sum=sum(n)) 

reasons<-reasons %>% 
  mutate(percent=(sum/sum(reasons$sum))*1000) 
reasons$percent[2]<-1
print(reasons)

# simple bar chart
ggplot(reasons,aes(x = main_reason, y = sum)) + 
  geom_bar(stat="identity")+
  coord_flip()+
  theme_minimal_hgrid(10, rel_small = 1)

pal9<-c(wes_palette("Royal1"), wes_palette("Moonrise3"))
#show_col(pal9)
ggplot(reasons, aes(fill = main_reason, values = percent))+
  geom_waffle(n_rows = 25, size = 0.33, colour = "white", flip = TRUE) +
  coord_equal() +
  theme_minimal_grid(10, rel_small = 1,line_size=0)+
  theme_enhance_waffle()+
  guides(fill = guide_legend(reverse = TRUE))+
  ggtitle("   Main reason for participation")+
  scale_fill_manual(
    name = NULL,
    values = c(pal9[5],pal9[2],pal9[7],pal9[4],pal9[1],pal9[6]),
    labels = c('For Donation',  'For Research',  'For Storing Eggs',
               'For Storing Embryos' ,'Reason unassigned' ,'Treatment Now' )
  ) 
```



```{r}
# update dat with clean categories
full_data<-full_data %>% 
  separate(`Main Reason for Producing Embroys Storing Eggs`, 
           into=c("main_reason"),sep=",", remove = F) %>%
  mutate(main_reason=str_trim(main_reason, side="right") )%>%  
  mutate(main_reason=ifelse(is.na(main_reason), "Reason unassigned", main_reason))

# we are only going to look at people who joined specifically for treatment, with known age
dat<-full_data %>% 
      filter(main_reason == 'Treatment Now') %>% 
      filter(`Patient Age at Treatment` != '999')
```



### Exploring types of infertility

Infertility is usually only diagnosed when a couple have not managed to conceive after a year of trying.

There are 2 types of infertility:

* **primary infertility** – where someone who's never conceived a child in the past has difficulty conceiving
* **secondary infertility **– where someone has had 1 or more pregnancies in the past, but is having difficulty conceiving again

*From: https://www.nhs.uk/conditions/infertility/*

```{r}
# get columsn for types of infertility
cols<-colnames(dat)[grepl("Type of Inf", colnames(dat))]
print(cols)
subset1<-dat %>% select(`Patient Age at Treatment`, cols) 
colnames(subset1)<-str_replace(colnames(subset1), "Type of Infertility -", "") %>%
                  str_trim(., side="both") 

print(subset1[398:406,2:7])

# convert dat to long format 
subset1_tidy<- subset1 %>% 
          pivot_longer(2:7, 
                    names_to = "type_of_infertility",
                    values_to = "binary") %>% 
              filter(binary == 1) %>% 
             # add partner categories
              mutate(partner=ifelse(grepl("Female", type_of_infertility), "Female",
                              ifelse(grepl("Male", type_of_infertility), "Male",
                              ifelse(grepl("Couple", type_of_infertility), "Couple",0))))
head(subset1_tidy)
```

```{r}
pal <- c(wes_palette("Moonrise3") , wes_palette("Royal1"))
# most basic
ggplot(subset1_tidy,aes(x = `Patient Age at Treatment`, fill = type_of_infertility)) + 
  geom_bar()+
  theme_minimal_hgrid(10, rel_small = 1)+
  scale_fill_manual(values=pal)+
  labs(fill = "Types of Infertility", y="")    

# by partner
ggplot(subset1_tidy,aes(x = `Patient Age at Treatment`, fill = type_of_infertility)) + 
  geom_bar()+
  theme_minimal_hgrid(10, rel_small = 1)+
  facet_wrap(~partner, nrow=1)+
  scale_fill_manual(values=pal)+
  labs(fill = "Types of Infertility", y="") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
# then I decided to convert each group into percentage

# create counts by group 
subset1_tidy2<- subset1_tidy %>% 
                group_by(`Patient Age at Treatment`, type_of_infertility) %>%         
                mutate(count = n()) %>%
                distinct() %>% 
                ungroup()
head(subset1_tidy2)

# calulate sums per age
age_counts<-subset1_tidy2 %>%
  select(`Patient Age at Treatment`, count) %>% 
  group_by(`Patient Age at Treatment`) %>% 
  summarize(total = sum(count)) %>%
  ungroup() %>% 
  column_to_rownames("Patient Age at Treatment") %>% 
  t() %>% as.data.frame()
print(age_counts)  

subset1_tidy2<-subset1_tidy2 %>% 
    mutate(age_group_percent= ifelse(`Patient Age at Treatment` == "18-34", count/age_counts$`18-34`,
                              ifelse(`Patient Age at Treatment` == "35-37", count/age_counts$`35-37`,
                              ifelse(`Patient Age at Treatment` == "38-39", count/age_counts$`38-39`,
                              ifelse(`Patient Age at Treatment` == "40-42", count/age_counts$`40-42`,
                              ifelse(`Patient Age at Treatment` == "43-44", count/age_counts$`43-44`,
                              ifelse(`Patient Age at Treatment` == "45-50", count/age_counts$`45-50`, NA)))))))
head(subset1_tidy2)

# now can definitely see the conncetion of age + primary/secondary 
pal <- c(wes_palette("Moonrise3") , wes_palette("Royal1"))

# 100 % by age group 
ggplot(subset1_tidy2,aes(x = `Patient Age at Treatment`, y=age_group_percent, fill = type_of_infertility)) + 
  geom_bar(stat = "identity")+
  theme_minimal_hgrid(10, rel_small = 1)+
  scale_fill_manual(values=pal)+
  #facet_wrap(~partner, nrow=1)+
  labs(fill = "Types of Infertility", y="")+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# stratigy by gender
ggplot(subset1_tidy2,aes(x = `Patient Age at Treatment`, y=age_group_percent, fill = type_of_infertility)) + 
  geom_bar(stat = "identity")+
  theme_minimal_hgrid(10, rel_small = 1)+
  scale_fill_manual(values=pal)+
  facet_wrap(~partner, nrow=1)+
  labs(fill = "Types of Infertility", y="")+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
# anyway now going to focus on male and female infertiliry only  

subset1_grouped <- subset1 %>% 
                select(1:5) %>% # keep female and male
                unite("types", 2:5,  sep="/" ) %>%  # group them into one column of binary 0/1
                group_by(`Patient Age at Treatment`, types) %>%         
                mutate(count = n()) %>%
                distinct() %>% ungroup()

colnames(subset1)[2:5]
head(subset1_grouped)


pal_cont <- wes_palette("Moonrise3", 17, type = "continuous")
#show_col(pal_cont)
ggplot(subset1_grouped,aes(x = `Patient Age at Treatment`, y= count, fill = types)) + 
  geom_bar(stat = "identity")+
  theme_minimal_hgrid(10, rel_small = 1)+
  facet_wrap(~types, nrow=10)+
  theme(legend.position = "none")+
  scale_fill_manual(values=pal_cont)+
  labs(fill = "Combination of infertily factors", y="")+
  guides(fill=guide_legend(ncol=1))

# count by type
subset1_grouped_counts<-subset1_grouped %>% 
              group_by( types) %>% 
              summarize(total = sum(count)) %>%
              ungroup() %>% 
              arrange(desc(total))

# look at most common types closer  
types_sub<-subset1_grouped_counts %>% filter(total >40000) %>% pull(types)
subset1_grouped %>% filter(types %in% types_sub)  %>% 
  ggplot(aes(x = `Patient Age at Treatment`, y= count, fill = types)) + 
    geom_bar(stat = "identity")+
    theme_minimal_hgrid(10, rel_small = 1)+
    facet_wrap(~types, nrow=5)+
    scale_fill_manual(values=pal_cont)+
    theme(legend.position = "none")+
    labs(fill = "Combination of infertily factors", y="")+
    guides(fill=guide_legend(ncol=1))

# look at the rest closer
types_sub<-subset1_grouped_counts %>% filter(total <40000) %>% pull(types)
subset1_grouped %>% filter(types %in% types_sub)  %>% 
  ggplot(aes(x = `Patient Age at Treatment`, y= count, fill = types)) + 
    geom_bar(stat = "identity")+
    theme_minimal_hgrid(10, rel_small = 1)+
    facet_wrap(~types, nrow=5)+
    theme(legend.position = "none")+
    scale_fill_manual(values=rev(pal_cont))+
    labs(fill = "Combination of infertily factors", y="")+
    guides(fill=guide_legend(ncol=1))+
    theme(axis.text.x = element_text(angle = 45, hjust = 1))



#subset1_grouped2 <- subset1 %>% select(1:5) %>% 
#                mutate(female_infertility= ifelse(rowSums(.[2:3]) == 0 , 0 , 1 )) %>% 
#                mutate(male_infertility= ifelse(rowSums(.[4:5]) == 0 , 0 , 1 )) %>% 
#                select(1,6,7) %>% 
#                unite("types", 2:3,  sep="/" ) %>% 
#                group_by(`Patient Age at Treatment`, types) %>%         
#                mutate(count = n()) %>%
#                distinct()
#                  
#
#pal <- c(wes_palette("Moonrise3") , wes_palette("Royal1"))
#pal <- c(wes_palette("Moonrise3")[1:2], wes_palette("Moonrise3")[1:2], wes_palette("Moonrise3")[1:2])
#ggplot(subset1_grouped2,aes(x = `Patient Age at Treatment`, y= count, fill = types)) + 
#  geom_bar(stat = "identity")+
#  theme_minimal_hgrid(10, rel_small = 1)+
#  facet_wrap(~types, nrow=10)+
#  #scale_fill_manual(values=pal)+
#  labs(fill = "Combination of infertily factors", y="")+
#  guides(fill=guide_legend(ncol=1))
```

### Exploring causes of infertility


```{r p, fig.height = 5, fig.width = 10}
# types of infertility
# get columsn for types of infertility
cols<-colnames(dat)[grepl("Cause", colnames(dat))]
print(cols)
subset2<-dat %>% select(`Patient Age at Treatment`, cols) %>% 
   rename("Cause of Infertility - Tubal disease"="Cause  of Infertility - Tubal disease",
         "Cause of Infertility - Partner Sperm Motility" = "Causes of Infertility - Partner Sperm Motility")
colnames(subset2)<-str_replace(colnames(subset2), "Cause of Infertility - ", "")%>%
                  str_trim(., side="both") 

print(subset2[398:406,2:7])



subset2<- subset2 %>% 
  pivot_longer(2:12, 
               names_to = "cause_of_infertility",
               values_to = "binary") %>%
  filter(binary == 1) %>% 
  mutate(partner=ifelse(grepl("Female", cause_of_infertility), "Female/Patient",
                 ifelse(grepl("Patient", cause_of_infertility), "Female/Patient",
                 ifelse(grepl("Male", cause_of_infertility), "Male/Partner",
                 ifelse(grepl("Partner", cause_of_infertility),"Male/Partner","Female/Patient"))))) %>% 
  mutate(cause_of_infertility=str_replace(cause_of_infertility,"Partner ", ""))

pal <- c(wes_palette("Royal1"), wes_palette("Moonrise3"), wes_palette("Chevalier1"))
p1<-subset2 %>% 
  filter(partner == "Female/Patient") %>% 
  ggplot(aes(x = `Patient Age at Treatment`, fill = cause_of_infertility)) + 
    geom_bar()+
    theme_minimal_hgrid(10, rel_small = 1)+
    scale_fill_manual(values=pal)+
    facet_wrap(~partner, nrow=1)+
    theme(legend.position = "right")+
    labs(fill = "Cause of Infertility", y="")+
    theme(axis.text.x = element_text(angle = 45, hjust = 1))

p2<-subset2 %>% 
  filter(partner == "Male/Partner") %>% 
  ggplot(aes(x = `Patient Age at Treatment`, fill = cause_of_infertility)) + 
    geom_bar()+
    theme_minimal_hgrid(10, rel_small = 1)+
    scale_fill_manual(values=rev(pal))+
    facet_wrap(~partner, nrow=1)+
    theme(legend.position = "right")+
    labs(fill = "Cause of Infertility", y="")+
    theme(axis.text.x = element_text(angle = 45, hjust = 1))

p<-(p1 + p2)
p
```



### Exploring types of treatments


```{r, echo=F}
pal17<-c("#d94b34",
         "#5dcf5c",
         "#b04bd7",
         "#a7da42",
         "#7666cb",
         "#d0bf4e",
         "#d04c99",
         "#7ed5a3",
         "#ba5a62",
         "#62883c",
         "#698fcb",
         "#b87b39",
         "#7dc2cd",
         "#d49fcb",
         "#607561",
         "#776685",
         "#cdb395")
```


```{r}
subset3 <- full_data %>% 
  select("Patient Age at Treatment", 
         "Year of Treatment", "years",
          "Type of treatment - IVF or DI" , "Specific treatment type", 
          "Egg Source" , "Sperm From" ,
          "Donated embryo" , "Patient acting as Surrogate") 

subset3 %>% count(`Type of treatment - IVF or DI`)
subset3 %>% count(`Specific treatment type`) %>%  arrange(desc(n))
```


* **IVF (In Vitro Fertilisation)** - An egg being fertilised by sperm outside the body.

* **DI (Donor Insemination)** - Using sperm from a sperm donor in order to get pregnant. This can be used in IUI, but also IVF and other fertility treatment options.

* **ICSI (Intracytoplasmic Sperm Injection)** - When a lone, high-quality sperm is injected straight into your egg during IVF, rather than allowing the sperm and egg to find one another in the dish.


```{r}
subset3 %>% pull(`Specific treatment type`) %>% unique()

subset3_tidy <-  subset3 %>% 
  mutate(`Specific treatment type` = str_replace_all(`Specific treatment type`, " / ", ":")) %>% 
  mutate(`Specific treatment type` = str_replace(`Specific treatment type`, "\\+", ":")) %>% 
  mutate(`Specific treatment type` = str_replace_all(`Specific treatment type`, ":Unknown", "")) %>% 
  mutate(`Specific treatment type` = str_replace_all(`Specific treatment type`, " ", "")) %>% 
  mutate(`Specific treatment type` = ifelse(is.na(`Specific treatment type`), "Unknown", `Specific treatment type`)) %>% 
  separate(`Specific treatment type`, into=c("main_treatment"), sep=":", remove=F) 

subset3_tidy %>% pull(main_treatment) %>% unique()
```

```{r}
# two main types of IVF
pal <- c(wes_palette("Royal1"), wes_palette("Moonrise3"), wes_palette("Darjeeling2"))
ggplot(subset3,aes(x = years, fill = `Type of treatment - IVF or DI`)) + 
  geom_bar()+
  theme_minimal_hgrid(10, rel_small = 1)+
  scale_fill_manual(values=pal)

# all IVF treatments by year
subset3_tidy %>% 
  filter(`Type of treatment - IVF or DI` == 'IVF') %>% 
ggplot(aes(x = years, fill = main_treatment)) + 
  geom_bar()+
  theme_minimal_hgrid(10, rel_small = 1)+
  scale_fill_manual(values=pal)


# now look only at rare treatments
subset3_tidy_sub<-subset3_tidy %>% 
  group_by(main_treatment) %>%
  filter(n()<5000) %>% 
  filter(n()>20)

ggplot(subset3_tidy_sub, aes(x = years, fill = main_treatment)) + 
  geom_bar()+
  theme_minimal_hgrid(10, rel_small = 1)+
  scale_fill_manual(values=pal)
```


```{r}
# looking at DI treatments
subset3_DI<-subset3_tidy %>% 
  filter(`Type of treatment - IVF or DI` == "DI") %>%
  select(main_treatment, years) %>%
  group_by(main_treatment, years) %>%
  mutate(count = n())%>%
  distinct() %>% 
  arrange(years)

pal <- c(wes_palette("Darjeeling2"))
ggplot(subset3_DI,aes(x = years, y= count,fill = main_treatment)) + 
  geom_col()+
  theme_minimal_hgrid(10, rel_small = 1)+
  scale_fill_manual(values=pal)

```



```{r, pa,  fig.height = 5, fig.width = 10 }
# finally, looking at treatment by specific years

dat<-subset3_tidy %>% 
  group_by(main_treatment) %>%
  #filter(n()<5000) %>% 
  #filter(n()>20) %>% 
  ungroup %>% 
  select(main_treatment, `Year of Treatment` )  %>% 
  group_by(main_treatment, `Year of Treatment`) %>% 
  mutate(count = n())%>%
  distinct() 

pal<-c(unname(piratepal("basel")), unname(piratepal("bugs")))
pa<- ggplot(data = dat, aes(x = `Year of Treatment`  , y = count, group = main_treatment)) +
  geom_line(aes(color = main_treatment, alpha = 1), size = 1) +
  geom_point(aes(color = main_treatment, alpha = 1), size = 3) +
  scale_x_continuous(breaks = sort(unique(dat$`Year of Treatment`))[c(TRUE, FALSE)]  )+
  theme(legend.position = "right") +
  scale_colour_manual(values=pal)+
  theme_minimal_hgrid(10, rel_small = 1) +
  labs(x = "year",  colour="Main treatment",
       y = "treatment",
       title = "Choice of treatments over the years (all)") +
  guides(alpha = FALSE) 
pa
```

```{r pb,  fig.height = 5, fig.width = 7 }

dat2<-subset3_tidy %>% 
  group_by(main_treatment) %>%
  filter(n()<5000) %>% 
  #filter(n()>20) %>% 
  ungroup %>% 
  select(main_treatment, `Year of Treatment` )  %>% 
  group_by(main_treatment, `Year of Treatment`) %>% 
  mutate(count = n())%>%
  distinct() 

pb <-ggplot(data = dat2, aes(x = `Year of Treatment`  , y = count, group = main_treatment)) +
  geom_line(aes(color = main_treatment, alpha = 1), size = 1) +
  geom_point(aes(color = main_treatment, alpha = 1), size = 3) +
  scale_x_continuous(breaks = sort(unique(dat$`Year of Treatment`))[c(TRUE, FALSE)]  )+
  theme(legend.position = "right") +
  scale_colour_manual(values=c(pal))+
  theme_minimal_hgrid(10, rel_small = 1) +
  labs(x = "year",  colour="Main treatment",
       y = "treatment",
       title = "Choice of treatments over the years (rare)")+
    guides(alpha = FALSE) 
pb
```



