---
title: "R Notebook"
output: html_notebook
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning =  FALSE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(comment = "")
```

```{r}
library(tidyverse)
library(tidymodels)
library(vroom)
library(forcats)
```


```{r}
# local 
data_path<-"~/Documents/Projects/IVF/data/"
```

```{r}
raw_data <-vroom(paste0(data_path, "hfea-ivf.csv"))
full_data <-raw_data %>% 
            # tidy age groups
            mutate(patient_age = gsub(" ", "", `Patient Age at Treatment`) ) %>% 
            # update year cols
            mutate(cohort=`source` %>% 
                  stringr::str_match("data/raw_data/ar-(.+?)-xlsb") %>% `[`(,2))

# update dat with clean categories
full_data <-full_data %>% 
  separate(`Main Reason for Producing Embroys Storing Eggs`, 
           into=c("main_reason"),sep=",", remove = F) %>%
  mutate(main_reason=str_trim(main_reason, side="right") )%>%  
  mutate(main_reason=ifelse(is.na(main_reason), "Reason unassigned", main_reason))

# we are only going to look at people who joined specifically for treatment, with known age, iafter 2000
dat1<-full_data %>% 
      filter(main_reason != 'For Research') %>%  # possibly include other categories later too
      filter(`Patient Age at Treatment` != '999') %>% 
      filter(!cohort %in% c('1991-1994', '1995-1999' ))

```

```{r}
# this shows that all DI treatments were performed in "Reason unassigned" enrolment type,
# so it's a perfect predicted, and therefore has to be excluded
full_data %>% filter(main_reason == "Reason unassigned") %>% count(`Type of treatment - IVF or DI` )
full_data %>% filter(main_reason != "Reason unassigned") %>% count(`Type of treatment - IVF or DI` )
```


```{r}
# tidy up ovulation induction methods
dat<-dat1 %>%
  rename(ovulation_induction=`Type of Ovulation Induction`) %>% 
  mutate(ovulation_induction= str_replace_all(ovulation_induction, " ,", ":"))  %>% 
  mutate(ovulation_induction= str_replace_all(ovulation_induction, "-", ""))  %>% 
  mutate(ovulation_induction= ifelse(ovulation_induction =="Yes but not recorded", "Unknown method",
                              ifelse(ovulation_induction == "Not stated but extrapolated",  "Unknown method",
                              ifelse(ovulation_induction=="Natural", "None", ovulation_induction)))) %>% 
  mutate(ovulation_induction = fct_lump_min(ovulation_induction %>% as.factor, min=50)) %>% 
  mutate_if(is.factor, as.character)
  
dim(dat[is.na(dat$ovulation_induction),])
dat[is.na(dat$ovulation_induction),]$ovulation_induction <- "No record"
dat %>% count(ovulation_induction) %>% View()
```




```{r}
# tidy up treatment subclasses
dat<-  dat %>%    
  rename(specific_treatment=`Specific treatment type`) %>% 
  mutate(specific_treatment = str_replace_all(specific_treatment, " / ", ":"),
         specific_treatment = str_replace(specific_treatment, "\\+", ":"),
         specific_treatment = str_replace_all(specific_treatment, ":Unknown", ""),
         specific_treatment = str_replace_all(specific_treatment, " ", ""),
         specific_treatment = ifelse(is.na(specific_treatment), "Unknown", specific_treatment)) %>% 
  separate(specific_treatment, into=c("specific_treatment"), sep=":", remove=F) %>% 
  mutate(specific_treatment = fct_lump_min(specific_treatment %>% as.factor, min=300)) %>% 
  #count(main_treatment) %>% View()
  filter(specific_treatment %in% c("IVF", "ICSI", "IUI", "ICI"))

dim(dat)
```


```{r}
dat_sub<-dat %>% 
    select(
        patient_age=`Patient Age at Treatment`,
        #duration_infertility=`Date patient started trying to become pregnant OR date of last pregnancy`, # 96% missing
        ##main_reason, # linaerly correlated with outcome
        cycles_IVF=`Total Number of Previous IVF cycles`,                                    
        cycles_DI=`Total Number of Previous DI cycles`,                                     
        pregnancy_IVF=`Total number of IVF pregnancies`,                                      
        pregnancy_DI=`Total number of DI pregnancies`,                                        
        birth_IVF=`Total number of live births - conceived through IVF`,                   
        birth_DI= `Total number of live births - conceived through DI`,
        infertility_FP =`Type of Infertility - Female Primary`,
        infertility_FS =`Type of Infertility - Female Secondary`,
        infertility_MP =`Type of Infertility - Male Primary`,
        infertility_MS =`Type of Infertility - Male Secondary`,
        infertility_CP =`Type of Infertility -Couple Primary`,
        infertility_CS =`Type of Infertility -Couple Secondary`,
        cause_tubaldisease = `Cause  of Infertility - Tubal disease`,                    
        cause_ovulatory = `Cause of Infertility - Ovulatory Disorder`   ,              
        cause_malefactor = `Cause of Infertility - Male Factor` ,                        
        cause_patient = `Cause of Infertility - Patient Unexplained`,               
        cause_endometrosis = `Cause of Infertility - Endometriosis` ,                      
        cause_cervical = `Cause of Infertility - Cervical factors`,                   
        cause_femalefactor = `Cause of Infertility - Female Factors` ,                    
        cause_sperm_conc = `Cause of Infertility - Partner Sperm Concentration`,        
        cause_sperm_morph = `Cause of Infertility -  Partner Sperm Morphology` ,          
        cause_sperm_motility = `Causes of Infertility - Partner Sperm Motility` ,            
        cause_sperm_immuno = `Cause of Infertility -  Partner Sperm Immunological factors`,
        #egg_source=`Egg Source`,
        #sperm_source=`Sperm From`,
        ovulation_induction,
        main_treatment=`Type of treatment - IVF or DI`) %>% 
 # mutate(egg_source=ifelse(is.na(egg_source), "unknown",egg_source)) %>% 
  mutate(cycles_IVF = ifelse(cycles_IVF == ">=5", 10 ,cycles_IVF)) %>%  # arbitrarily going to give this a large value
  mutate(cycles_DI = ifelse(cycles_DI == ">=5", 10 ,cycles_DI)) 

dat_sub<-dat_sub %>% 
          mutate_if(is.character, as.factor) %>% 
          mutate_if(is.numeric, as.factor) %>% 
          mutate_at(vars(
                     cycles_IVF,
                     cycles_DI,
                     pregnancy_IVF,
                     pregnancy_DI,
                     birth_IVF,
                     birth_DI), as.numeric) %>% 
          filter_all(all_vars(!is.na(.)))
```


## Predict treatment IVF vs DI
```{r}
# split into train/test
data_split <- initial_split(dat_sub, prop = 0.6)

# preprocess
my_recipe <- training(data_split) %>%
              recipe(main_treatment ~.) %>%
              step_dummy(all_predictors()) %>% 
              step_corr(all_predictors()) %>%
              prep()
# apply recipe to train and test
my_testing <- my_recipe %>%
               bake(testing(data_split)) 
my_training <- juice(my_recipe)
```


```{r}
# model 1
rf <-  rand_forest(trees = 20, mode = "classification") %>%
  set_engine("randomForest") %>%
  fit(main_treatment ~ ., data = my_training)

out1<-rf %>%
  predict(my_testing) %>%
  bind_cols(my_testing)

out1 %>% 
metrics(truth=main_treatment, estimate=.pred_class)


out1_prob <- rf %>%
  predict(my_testing, type = "prob") %>%
  bind_cols(my_testing)

out1_prob %>%
    mutate_at(vars(.pred_IVF), function(x) as.integer(as.character(x))) %>%
    roc_auc(., main_treatment, .pred_IVF)

out1_prob %>% 
  roc_curve(truth=main_treatment, .pred_IVF) %>%
  autoplot()


vip::vip(rf)
```


```{}
#Based on age+ infertility + numbers + ovulation induction
  .metric  .estimator .estimate
  <chr>    <chr>          <dbl>
1 accuracy binary         0.980
2 kap      binary         0.881
  roc_auc	 binary	    0.9484991	
  
  
#Based on infertility + numbers
  .metric  .estimator .estimate
  <chr>    <chr>          <dbl>
1 accuracy binary         0.962
2 kap      binary         0.762


# based on infertility + numbers (except cycles)

  .metric  .estimator .estimate
  <chr>    <chr>          <dbl>
1 accuracy binary         0.924
2 kap      binary         0.430
  roc_auc	 binary	        0.33


# based on age + infertility
.metric  .estimator .estimate
  <chr>    <chr>          <dbl>
1 accuracy binary         0.919
2 kap      binary         0.276
  roc_auc	 binary	        0.40

```
Kappa or Cohen’s Kappa is like classification accuracy, except that it is normalized at the baseline of random chance on your dataset. It is a more useful measure to use on problems that have an imbalance in the classes (e.g. 70-30 split for classes 0 and 1 and you can achieve 70% accuracy by predicting all instances are for class 0). 


```{r}
# model 2
lg <-  logistic_reg(penalty=0) %>%
  set_engine("glmnet") %>%
  fit(main_treatment ~ ., data = my_training)

out2<-lg %>%
  predict(my_testing) %>%
  bind_cols(my_testing)

out2 %>% 
metrics(truth=main_treatment, estimate=.pred_class)
```

```{}
# based on age + infertility
  .metric  .estimator .estimate
  <chr>    <chr>          <dbl>
1 accuracy binary         0.914
2 kap      binary         0.226

#Based on infertility + numbers

  .metric  .estimator .estimate
  <chr>    <chr>          <dbl>
1 accuracy binary         0.943
2 kap      binary         0.608
```


```{r}
# model 3
xg <-  boost_tree(trees=100,  mode = "classification") %>%
  set_engine("xgboost") %>%
  fit(main_treatment ~ ., data = my_training)

out3<-xg %>%
  predict(my_testing) %>%
  bind_cols(my_testing)

out3 %>% 
metrics(truth=main_treatment, estimate=.pred_class)
```


```{r}
# Based on infertility + numbers

 .metric  .estimator .estimate
  <chr>    <chr>          <dbl>
1 accuracy binary         0.962
2 kap      binary         0.761
```




## Sub-categories of treatments
```{r}

dat_sub<-dat %>% 
    select(
        patient_age=`Patient Age at Treatment`,
        #cycles_IVF=`Total Number of Previous IVF cycles`,                                    
        #cycles_DI=`Total Number of Previous DI cycles`,                                     
        #pregnancy_IVF=`Total number of IVF pregnancies`,                                      
        #pregnancy_DI=`Total number of DI pregnancies`,                                        
        #birth_IVF=`Total number of live births - conceived through IVF`,                   
        #birth_DI= `Total number of live births - conceived through DI`,
        infertility_FP =`Type of Infertility - Female Primary`,
        infertility_FS =`Type of Infertility - Female Secondary`,
        infertility_MP =`Type of Infertility - Male Primary`,
        infertility_MS =`Type of Infertility - Male Secondary`,
        infertility_CP =`Type of Infertility -Couple Primary`,
        infertility_CS =`Type of Infertility -Couple Secondary`,
        cause_tubaldisease = `Cause  of Infertility - Tubal disease`,                    
        cause_ovulatory = `Cause of Infertility - Ovulatory Disorder`   ,              
        cause_malefactor = `Cause of Infertility - Male Factor` ,                        
        cause_patient = `Cause of Infertility - Patient Unexplained`,               
        cause_endometrosis = `Cause of Infertility - Endometriosis` ,                      
        cause_cervical = `Cause of Infertility - Cervical factors`,                   
        cause_femalefactor = `Cause of Infertility - Female Factors` ,                    
        cause_sperm_conc = `Cause of Infertility - Partner Sperm Concentration`,        
        cause_sperm_morph = `Cause of Infertility -  Partner Sperm Morphology` ,          
        cause_sperm_motility = `Causes of Infertility - Partner Sperm Motility` ,            
        cause_sperm_immuno = `Cause of Infertility -  Partner Sperm Immunological factors`,
        #ovulation_induction,
        specific_treatment) %>%
  mutate(specific_treatment=droplevels(specific_treatment)) %>% 
 # mutate(egg_source=ifelse(is.na(egg_source), "unknown",egg_source)) %>% 
  mutate(cycles_IVF = ifelse(cycles_IVF == ">=5", 10 ,cycles_IVF)) %>%  # arbitrarily going to give this a large value
  mutate(cycles_DI = ifelse(cycles_DI == ">=5", 10 ,cycles_DI)) 

dat_sub<-dat_sub %>% 
          mutate_if(is.character, as.factor) %>% 
          mutate_if(is.numeric, as.factor) %>% 
          #mutate_at(vars(
          #            cycles_IVF,
          #            cycles_DI,
          #            pregnancy_IVF,
          #            pregnancy_DI,
          #            birth_IVF,
          #            birth_DI), as.numeric) %>% 
          filter_all(all_vars(!is.na(.)))
```

```{r}
# split into train/test
data_split <- initial_split(dat_sub, prop = 0.6)

# preprocess
my_recipe <- training(data_split) %>%
              recipe(specific_treatment ~.) %>%
              step_dummy(all_predictors()) %>% 
              step_corr(all_predictors()) %>%
              prep()
# apply recipe to train and test
my_testing <- my_recipe %>%
               bake(testing(data_split)) 
my_training <- juice(my_recipe)
```


```{r}
# model 1
rf_multi <-  rand_forest(trees = 50, mode = "classification") %>%
  set_engine("randomForest") %>%
  fit(specific_treatment ~ ., data = my_training)

out1_multi<-rf_multi %>%
  predict(my_testing) %>%
  bind_cols(my_testing)

out1_multi %>% 
metrics(truth=specific_treatment, estimate=.pred_class)


out1_multi %>% select(specific_treatment, .pred_class) %>% View()


predict(rf_multi, my_testing, type = "prob") %>%
  bind_cols(predict(rf_multi, my_testing)) %>%
  bind_cols(select(my_testing, specific_treatment)) %>%View()

  metrics(specific_treatment, .pred_ICI:.pred_IVF, estimate = .pred_class)
View(class_voting)



out1_multi_prob <- rf_multi %>%
  predict(my_testing, type = "prob") %>%
  bind_cols(my_testing)

out1_multi_prob %>%
    mutate_at(vars(.pred_IVF), function(x) as.integer(as.character(x))) %>%
    roc_auc(., specific_treatment, .pred_ICI:.pred_IVF)


out1_multi_prob %>% 
  roc_curve(truth=specific_treatment, .pred_ICI:.pred_IVF) %>%
  autoplot()


vip::vip(rf_multi)
```
 
```{}
 #Based on infertility + numbers
  .metric  .estimator .estimate
  <chr>    <chr>          <dbl>
1 accuracy multiclass     0.726
2 kap      multiclass     0.530

  .metric .estimator .estimate
  <chr>   <chr>          <dbl>
1 roc_auc hand_till      0.730

 #Based on infertility + numbers + induction
 .metric  .estimator .estimate
  <chr>    <chr>          <dbl>
1 accuracy multiclass     0.744
2 kap      multiclass     0.566
  .metric .estimator .estimate
  <chr>   <chr>          <dbl>
1 roc_auc hand_till      0.755

# base only on age+ infertility
  .metric  .estimator .estimate
  <chr>    <chr>          <dbl>
1 accuracy multiclass     0.678
2 kap      multiclass     0.423
roc_auc	hand_till	0.6047284	
```

```{r}
# model 2
knn_multi <-  nearest_neighbor(mode = "classification") %>%
  set_engine("kknn") %>%
  fit(specific_treatment ~ ., data = my_training)

out1_multi<-rf_multi %>%
  predict(my_testing) %>%
  bind_cols(my_testing)

out1_multi %>% 
metrics(truth=specific_treatment, estimate=.pred_class)
```
